Generating Correlated Data for Omics Simulation

### 2025-06-20 manuscript/paper.qmd
Revise according to editorial requests

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    margin-left: 0.75in
    margin-right: 0.75in
    margin-top: 0.75in
    margin-bottom: 0.75in
bibliography: references.bib
---
## Abstract

@@

```{r}
#| label: fig-compare-to-real
#| fig-cap: "Comparison to real reference data run on a mouse cortex dataset from GSE151923. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 100 reads were used. Values on the diagonal line indicate a match between the simulated and real datasets. (d) Projections onto the top two principal components of the real dataset for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all datasets and the variance captured by the top components is shown. Unlike (d), these components were fit from each dataset considered separately instead of reusing the weights from the real data."
#| fig-width: 10
#| fig-height: 10
readRDS("../processed/compare_to_real_plot.GSE151923.RDS")
```

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
Metabolomics data was obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).

@@

The authors declare no competing interests.

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania, Philadelphia, Pennsylvania, United States of America"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    margin-left: 0.75in
    margin-right: 0.75in
    margin-top: 0.75in
    margin-bottom: 0.75in
bibliography: references.bib
---
## Abstract

@@

```{r}
#| label: fig-compare-to-real
#| fig-cap: "Comparison to real reference data run on a mouse cortex dataset from GSE151923. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least 100 reads were used. Values on the diagonal line indicate a match between the simulated and real datasets. (d) Projections onto the top two principal components of the real dataset for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all datasets and the variance captured by the top components is shown. Unlike (d), these components were fit from each dataset considered separately instead of reusing the weights from the real data."
#| fig-width: 10
#| fig-height: 10
readRDS("../processed/compare_to_real_plot.GSE151923.RDS")
```

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
Metabolomics data was obtained from MetaboAnalyst (https://api2.xialab.ca/api/download/metaboanalyst/plasma_nmr.csv).

@@

The authors have declared that no competing interests exist.

## References

::: {#refs}
:::

## Supporting Information

### Supporting Information 1

Supporting document containing all Supplemental Figures 1-3 and Supplemental Methods.
